---
title: "index.qmd"
author: "DAGE"
format: html
editor: visual
---

# Neural Network Classification of Open-Text Survey Responses

# Introduction

The goal of this project was to determine if neural network models could successfully predict a person's survey outcomes using solely their responses from open-text questions. Two separate models were developed, each predicting a unique survey outcome from the respective open-text question used as inputs. The two models are:

1.  Predicting 'dominant transportation method' from the open-text responses to question:

    |  |
    |----|
    | "Describe your ideal way to get around your city, and why you prefer it. To help make your response has richly educational as possible for our students we request you consider your response to offer a thoughtful and meaningful entry." |

2.  Predicting 'dominant apprehension type' from the open-text responses to question:

    |  |
    |----|
    | "If a friend told you they get nervous talking to new people, what advice would you give them? To help make your response has richly educational as possible for our students we request you consider your response to offer a thoughtful and meaningful entry." |

## Defining Outcome Variables

The outcome variables for both of our models were engineered to enable each model to attempt a binary classification. This section will detail the process for creating these categories and assigning classes to all respondents.

### Dominant Transportation Method

Dominant Transportation is an engineered outcome variable that identifies the greatest of two interactive scores (if any): Estimated monthly rideshare usage & estimated monthly public transit usage.

```{r}
#| eval: FALSE
#| echo: TRUE
# Set numeric values for transportation variables
usage_monthly <- c(
  "Never"                     = 0,
  "0-1 days a month"          = 1,
  "2-4 days a month"          = 3,
  "4-8 days a month"          = 6,
  "8 or more days a month"    = 8
  )
usage_daily <- c(
  "1-2 rides in a typical day"       = 1.5,
  "3-4 rides in a typical day"       = 3.5,
  "5-6 rides in a typical day"       = 5.5,
  "7 or more rides in a typical day" = 7
  )

# Calculate highest usage transport method
survey <- survey |>
  mutate(
    
#   Convert 3-month usage variables to numeric
    across(c(q26_public_transit_3months, q28_rideshare_3months), ~ usage_monthly[.x]),
    
#   Convert typical daily usage variables to numeric
    across(c(q27_public_transit_dayuse, q29_rideshare_dayuse), ~ usage_daily[.x]),

#   Multiply usage over past 3 months by typical daily usage (usage count estimate)
    public_transit_count = q26_public_transit_3months * q27_public_transit_dayuse,
    rideshare_count = q28_rideshare_3months * q29_rideshare_dayuse,

#   Assign highest usage transport method (equal if BA)
    highest_transport = case_when(
      public_transit_count > rideshare_count ~ "public_transit",
      rideshare_count > public_transit_count ~ "rideshare",
      public_transit_count == 0 & rideshare_count == 0 ~ "neither",
      .default = "equal" )
    )
```

After numericizing values from a participants' survey responses, relevant questions were sorted into these two groups, depending on which transportation method each question targeted. Each group then had the corresponding values multiplied to estimate the number of monthly trips for both transportation methods. The Dominant Transportation Method was then assigned one of four values.

- If both scores have a value of 0, a participant is assigned to the 'neither' category (indicated no rideshare or public transit use)
- If both scores have a value \> 0, but are still equal, a participant is assigned to the 'equal' category (indicated equal usage of both methods)
- If there was a greater value between estimated rideshare usage and estimated public transit usage, a participant was assigned to the category with higher usage.

{{< embed ../../scripts/eda_dependents.qmd#transportation_grid >}}

The distribution of Dominant Transportation Methods was grossly imbalanced when considering all possible categories (rideshare, public transit, neither, equal), however the distribution became roughly balanced when collapsing categories down to public transit vs. all other possibilities combined. Our group elected to collapse categories as such in order to enable a binary classification approach using a relatively balanced sample.

*NOTE: The engineering of 'Dominant Transportation Method' occurs in the clean_data.qmd script.*

*NOTE: Rendering of the above figures occurs in the eda_dependents.qmd script.*

### Dominant Apprehension Type

Dominant Apprehension Type is an engineered outcome variable that identifies the greatest of two composite scores: Group Discussion Apprehension & Solo Conversation Apprehension.

```{r}
#| eval: FALSE
#| echo: TRUE
# Identify forward and reverse-coded Q's (Higher scores = more apprehensive)
forward_code <- c(
  "q1_dislike_groupdiscussion", "q3_nervous_groupdiscussion",
  "q5_tense_newppl_groupdiscussion", "q13_nervous_newppl_convo",
  "q15_nervous_convo", "q18_afraid_convo"
  )
reverse_code <- c(
  "q2_comfort_groupdiscussion", "q4_like_groupdiscussion",
  "q6_relaxed_groupdiscussion", "q14_nofear_speaking_convo",
  "q16_relaxed_convo", "q17_relaxed_newppl_convo"
  )
# Set numeric values for Likert scales
likert_forward <- c(
  "Strongly disagree"          = 1,
  "Somewhat disagree"          = 2,
  "Neither agree nor disagree" = 3,
  "Somewhat agree"             = 4,
  "Strongly agree"             = 5
)
likert_reverse <- c(
  "Strongly disagree"          = 5,
  "Somewhat disagree"          = 4,
  "Neither agree nor disagree" = 3,
  "Somewhat agree"             = 2,
  "Strongly agree"             = 1
)
# Convert to numeric
survey <- survey |>
  mutate(
    across(all_of(forward_code), ~ likert_forward[.x]),
    across(all_of(reverse_code), ~ likert_reverse[.x])
  )
# Identify group discussion vs. solo conversation Q's
group_discussion <- c(
  "q1_dislike_groupdiscussion", "q2_comfort_groupdiscussion",
  "q3_nervous_groupdiscussion", "q4_like_groupdiscussion",
  "q6_relaxed_groupdiscussion"
  )
solo_conversation <- c(
  "q13_nervous_newppl_convo", "q14_nofear_speaking_convo",
  "q15_nervous_convo",        "q16_relaxed_convo",
  "q17_relaxed_newppl_convo", "q18_afraid_convo"
  )
# Calculate biggest fear
survey <- survey |> 
  mutate(
    
#   Average of group apprehension Q's
    group_apprehension = rowMeans(
      pick(all_of(group_discussion)), na.rm = TRUE),
    
#   Average of solo apprehension Q's
    solo_apprehension = rowMeans(
      pick(all_of(solo_conversation)), na.rm = TRUE),

#   Assign greatest of the two values (equal becomes NA)
    biggest_fear = case_when(
      group_apprehension > solo_apprehension ~ "group_apprehension",
      solo_apprehension > group_apprehension ~ "solo_apprehension",
      .default = "equal")
  )
```

After numericizing Likert scale values from a participants PRCA responses, relevant questions were sorted into these two groups, depending on which type of apprehension each question targeted. Each group then had the corresponding numeric Likert scale values averaged to calculate the composite scores for 'average group discussion apprehension' and 'average solo conversation apprehension'. Finally, the Dominant Apprehension Type takes the largest of these two values and codes for the corresponding group, or indicates if both scores are equal.

{{< embed ../../scripts/eda_dependents.qmd#apprehension_grid >}}

The distribution of dominant apprehension types (either group discussions or solo conversations) was grossly imbalanced across all possible categories (group discussions, solo conversations, and equal apprehension) but becomes roughly balanced when excluding the equal apprehension group. Our group elected to exclude equal apprehension responses to enable a binary classification approach with a relatively balanced sample.

*NOTE: The engineering of 'Dominant Transportation Method' occurs in the clean_data.qmd script.*

*NOTE: Rendering of the above figures occurs in the eda_dependents.qmd script.*

## Tools

The following are a list of tools utilized throughout the project:

- R was used primarily for cleaning and EDA scripts prior to building the neural networks

- Python was utilized for building the neural networks and visualizing results, primarily relying on the TensorFlow, Keras, and Seaborn packages. Agentic IDEs including Devin and Google Colab supported this process.

- Quarto was utilized for creating research memos and the final manuscript.

- A GitHub repository was utilized by the team to manage collaboration and development history.

- A markdown file (contributions.md) was kept to record individual team member contributions to the project.

## Methods

Both models utilized similar neural network models to accomplish their tasks, with slight differences in model architectures and hyperparameter values that were individualized during model tuning.

With both models taking natural language data as inputs, the first step was to tokenize each vocabulary set, then vectorize participant responses accordingly. Both models tokenized on whole words, employed a maximum vocabulary size of 5000 tokens, and a maximum sequence length of 100 tokens. The vocabulary size limit means that only the 5000 most common words across the training set were given a unique token identifier, with any other rare words being assigned to a placeholder token value indicating that this word was not included in the vocabulary. The sequence length limit means that any responses longer than 100 words were truncated.

Each model then created a unique set of word embeddings with training data
